---
title: "Dataset Dictionary — Monday-6:30pm Group"
subtitle: "Intro to R for Data Science with LLMs — Workshop 2026"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
---

```{r}
library(tidyverse)
```

This document describes every CSV file in the `data/` folder: what one row
represents, and what each column means.  The datasets belong to four separate
projects:

1. **Group A — Yogurt intervention** (`data/group_a_yogurt/`)
2. **Group B — Menstruation study** (`data/group_b_menstruation/`)
3. **Instructional dataset — UKZN workshop 2023** (`data/instructional_dataset/`)
4. **Modeling — Sorghum time‑course** (`data/modeling/`)

---

# Group A — Yogurt intervention {#sec-yogurt}

## `00_sample_ids_yogurt.csv`

```{r}
sample_ids_yogurt <- read_csv("../data/group_a_yogurt/00_sample_ids_yogurt.csv")
glimpse(sample_ids_yogurt)
```

**One row =** one biological sample collected from a participant at one visit.

| Column       | Type     | Meaning |
|--------------|----------|---------|
| `pid`        | character | Participant ID (e.g. `pid_25`). Links to `01_participant_metadata_yogurt.csv`. |
| `time_point` | character | When the sample was taken: `baseline` (before any intervention) or `after_antibiotic` (after a course of antibiotics). |
| `arm`        | character | Study group: `yogurt` (received yogurt) or `unchanged_diet` (control). |
| `sample_id`  | character | Unique barcode for the lab sample (e.g. `UNCH001`, `YOG002`). Links to the qPCR and Luminex results. |

---

## `01_participant_metadata_yogurt.csv`

```{r}
meta_yogurt <- read_csv("../data/group_a_yogurt/01_participant_metadata_yogurt.csv")
glimpse(meta_yogurt)
```

**One row =** one participant enrolled in the yogurt study.

| Column              | Type     | Meaning |
|---------------------|----------|---------|
| `pid`               | character | Participant ID. Links to `00_sample_ids_yogurt.csv`. |
| `arm`               | character | Study arm: `yogurt` or `unchanged_diet`. |
| `days_since_last_sex` | numeric | Days since last sexual intercourse reported at enrolment. |
| `birth_control`     | character | Birth‑control method: `Depoprovera` (injectable), `no hormonal birth control`. |
| `age`               | numeric | Age in years. |
| `education`         | character | Highest education level: `less than grade 9` / `grade 10-12, not matriculated` / `post-secondary` / etc. |
| `sex`               | character | Self‑reported sex (`female` for all participants). |

---

## `02_qpcr_results_yogurt.csv`

```{r}
qpcr_yogurt <- read_csv("../data/group_a_yogurt/02_qpcr_results_yogurt.csv")
glimpse(qpcr_yogurt)
```

**One row =** qPCR measurements from one sample for three bacterial targets.

| Column             | Type     | Meaning |
|--------------------|----------|---------|
| `sample_id`        | character | Sample barcode. Links to `00_sample_ids_yogurt.csv`. |
| `qpcr_bacteria`    | numeric  | Total bacterial load (16S rRNA gene copies). |
| `qpcr_crispatus`   | numeric  | *Lactobacillus crispatus* — a beneficial vaginal species (gene copies). |
| `qpcr_iners`       | numeric  | *Lactobacillus iners* — another *Lactobacillus* species (gene copies). |

These numbers reflect the absolute abundance of each target. A value of `0`
means the target was not detected.

---

## `03_luminex_results_yogurt.csv`

```{r}
luminex_yogurt <- read_csv("../data/group_a_yogurt/03_luminex_results_yogurt.csv")
glimpse(luminex_yogurt)
```

**One row =** the concentration of one **cytokine** measured in one sample by
Luminex multiplex assay.

| Column      | Type     | Meaning |
|-------------|----------|---------|
| `sample_id` | character | Sample barcode. Links to `00_sample_ids_yogurt.csv`. |
| `cytokine`  | character | Name of the cytokine / chemokine measured (e.g. `IL-1a`, `IL-6`, `TNFa`, `IFN-Y`, `IP-10`, `MIG`, `MIP-3a`, `IL-8`, `IL-10`, `IL-1b`). |
| `conc`      | numeric  | Concentration (pg/mL). |
| `limits`    | character | Whether the value was `within limits` (reliable) or `out of range` (below/above the standard curve). |

Because the data is in **long format**, each sample appears in multiple rows —
one per cytokine.

---

# Group B — Menstruation study {#sec-menstruation}

## `00_sample_ids_period.csv`

```{r}
sample_ids_period <- read_csv("../data/group_b_menstruation/00_sample_ids_period.csv")
glimpse(sample_ids_period)
```

**One row =** one biological sample collected during the menstrual cycle study.

| Column       | Type     | Meaning |
|--------------|----------|---------|
| `pid`        | character | Participant ID (e.g. `pid_12`). Links to `01_participant_metadata_period.csv`. |
| `time_point` | character | Menstrual‑cycle visit: `week_prior`, `onset`, `end_bleeding`, `week_post`. |
| `arm`        | character | Study group: `birth_control` (using hormonal contraception) or `no_birth_control`. |
| `sample_id`  | character | Unique barcode for the lab sample (e.g. `SAMP001`). Links to Luminex and flow cytometry results. |

---

## `01_participant_metadata_period.csv`

```{r}
meta_period <- read_csv("../data/group_b_menstruation/01_participant_metadata_period.csv")
glimpse(meta_period)
```

**One row =** one participant enrolled in the menstruation study.

| Column           | Type     | Meaning |
|------------------|----------|---------|
| `pid`            | character | Participant ID. Links to `00_sample_ids_period.csv`. |
| `arm`            | character | `birth_control` or `no_birth_control`. |
| `age`            | numeric  | Age in years. |
| `pcos_status`    | character | Polycystic ovary syndrome: `pcos` or `no disease`. |
| `period_product` | character | Menstrual product used: `tampon`, `menstrual_cup`, or `pad`. |
| `sex`            | character | Self‑reported sex (`female`). |

---

## `02_luminex_period.csv`

```{r}
luminex_period <- read_csv("../data/group_b_menstruation/02_luminex_period.csv")
glimpse(luminex_period)
```

**One row =** the concentration of one **cytokine** measured in one sample.

| Column      | Type     | Meaning |
|-------------|----------|---------|
| `sample_id` | character | Sample barcode. Links to `00_sample_ids_period.csv`. |
| `cytokine`  | character | Name of the analyte: `IL-1a`, `IL-1b`, `IL-6`, `TNFa`, `IFNg`, `IP-10`, `MIG`, `MIP-3a`. |
| `conc`      | numeric  | Concentration (pg/mL, Luminex assay). |
| `limits`    | character | Detection status: `not_censored` (reliable), `below_detection_limit`, or `above_detect_limit`. |

Again, **long format** — one row per cytokine per sample.

---

## `03_flow_cytometry_period.csv`

```{r}
flow_period <- read_csv("../data/group_b_menstruation/03_flow_cytometry_period.csv")
glimpse(flow_period)
```

**One row =** the immune‑cell counts from one sample, measured by flow
cytometry.

| Column               | Type     | Meaning |
|----------------------|----------|---------|
| `sample_id`          | character | Sample barcode. Links to `00_sample_ids_period.csv`. |
| `live_cd19_negative` | numeric  | Total live cells minus CD19⁺ (B‑cell) events — a starting gate. |
| `cd45_negative`      | numeric  | Cells that are CD45⁻ (not leukocytes). |
| `cd45_positive`      | numeric  | Cells that are CD45⁺ (leukocytes / white blood cells). |
| `neutrophils`        | numeric  | Neutrophil count (a type of innate immune cell). |
| `non_neutrophils`    | numeric  | CD45⁺ cells that are not neutrophils. |
| `cd3_negative`       | numeric  | CD45⁺ cells that are CD3⁻ (not T‑cells). |
| `cd3_positive`       | numeric  | CD45⁺ cells that are CD3⁺ (T‑cells). |
| `cd4_t_cells`        | numeric  | CD4⁺ helper T‑cells (a subset of CD3⁺). |
| `cd8_t_cells`        | numeric  | CD8⁺ cytotoxic T‑cells (a subset of CD3⁺). |

All values are **absolute cell counts** (number of events).

---

# Instructional dataset — UKZN workshop 2023 {#sec-instructional}

## `00_sample_ids_UKZN_workshop_2023.csv`

```{r}
sample_ids_inst <- read_csv("../data/instructional_dataset/00_sample_ids_UKZN_workshop_2023.csv")
glimpse(sample_ids_inst)
```

**One row =** one biological sample from a vaginal microbiome study (original
UKZN 2023 workshop dataset).

| Column       | Type     | Meaning |
|--------------|----------|---------|
| `pid`        | character | Participant ID. Links to `01_participant_metadata_UKZN_workshop_2023.csv`. |
| `time_point` | character | Study visit: `baseline`, `week_1`, or `week_7`. |
| `arm`        | character | `placebo` (control) or `treatment`. |
| `sample_id`  | character | Lab barcode (e.g. `SAMP001`). Links to ELISA and flow results. |

---

## `01_participant_metadata_UKZN_workshop_2023.csv`

```{r}
meta_inst <- read_csv("../data/instructional_dataset/01_participant_metadata_UKZN_workshop_2023.csv")
glimpse(meta_inst)
```

**One row =** one participant.

| Column     | Type     | Meaning |
|------------|----------|---------|
| `pid`      | character | Participant ID. |
| `arm`      | character | `placebo` or `treatment`. |
| `smoker`   | character | Smoking status: `smoker` / `non-smoker`. |
| `age`      | numeric  | Age in years. |
| `education`| character | Education level: `less than grade 9`, `grade 10-12, matriculated`, `post-secondary`, etc. |
| `sex`      | character | `F` (female). |

---

## `02_visit_clinical_measurements_UKZN_workshop_2023.csv`

```{r}
clinical_inst <- read_csv("../data/instructional_dataset/02_visit_clinical_measurements_UKZN_workshop_2023.csv")
glimpse(clinical_inst)
```

**One row =** one clinical assessment from one participant at one visit.

| Column       | Type     | Meaning |
|--------------|----------|---------|
| `pid`        | character | Participant ID. |
| `time_point` | character | Visit: `baseline`, `week_1`, `week_7`. |
| `arm`        | character | `placebo` or `treatment`. |
| `nugent_score`| numeric | **Nugent score** (0–10) — a measure of vaginal microbiome health; higher = more disturbed. |
| `crp_blood`  | numeric  | C‑reactive protein in blood (mg/L) — a marker of systemic inflammation. |
| `ph`         | numeric  | Vaginal pH. |

---

## `03_elisa_cytokines_UKZN_workshop_2023.csv`

```{r}
elisa_inst <- read_csv("../data/instructional_dataset/03_elisa_cytokines_UKZN_workshop_2023.csv")
glimpse(elisa_inst)
```

**One row =** the concentration of one **cytokine** measured in one sample by
ELISA.

| Column      | Type     | Meaning |
|-------------|----------|---------|
| `sample_id` | character | Sample barcode. Links to `00_sample_ids_UKZN_workshop_2023.csv`. |
| `cytokine`  | character | Cytokine name (same panel as the Luminex datasets: `IL-1a`, `IL-6`, `TNFa`, `IFN-Y`, `IL-8`, `IL-10`, `IP-10`, `MIG`, `MIP-3a`, `IL-1b`). |
| `conc`      | numeric  | Concentration (pg/mL). |
| `limits`    | character | `within limits` (reliable) or `out of range` (below/above curve). |

Long format — one row per cytokine per sample.

---

## `04_flow_cytometry_UKZN_workshop_2023.csv`

```{r}
flow_inst <- read_csv("../data/instructional_dataset/04_flow_cytometry_UKZN_workshop_2023.csv")
glimpse(flow_inst)
```

**One row =** immune‑cell counts from one sample. Same flow‑cytometry panel as
the menstruation study.

| Column               | Type     | Meaning |
|----------------------|----------|---------|
| `sample_id`          | character | Sample barcode. |
| `live_cd19_negative` | numeric  | Total live cells minus CD19⁺ events (gating parent). |
| `cd45_negative`      | numeric  | CD45⁻ (non‑leukocyte) cells. |
| `cd45_positive`      | numeric  | CD45⁺ leukocytes. |
| `neutrophils`        | numeric  | Neutrophils (innate immune cells). |
| `non_neutrophils`    | numeric  | Leukocytes that are not neutrophils. |
| `cd3_negative`       | numeric  | Leukocytes that are CD3⁻ (not T‑cells). |
| `cd3_positive`       | numeric  | Leukocytes that are CD3⁺ (T‑cells). |
| `cd4_t_cells`        | numeric  | CD4⁺ helper T‑cells. |
| `cd8_t_cells`        | numeric  | CD8⁺ cytotoxic T‑cells. |

---

# Modeling — Sorghum time‑course {#sec-sorghum}

## `01-sorghum_timecourse.csv`

```{r}
sorghum <- read_csv("../data/modeling/01-sorghum_timecourse.csv")
glimpse(sorghum)
```

**One row =** one RNA‑seq measurement from one sorghum sample at one time
point.

| Column                     | Type      | Meaning |
|----------------------------|-----------|---------|
| `Sobic.001G062300.v3.1` …  | numeric   | Expression values (read counts or normalized counts) for 9 individual sorghum genes. Column names are gene IDs from the Sorghum bicolor genome v3.1. |
| `Barcode`                  | character | Sequencing library barcode (identifies the sample in the sequencing run). |
| `libraryName`              | character | Library name (e.g. `BOCAX`, `BOCGY`). |
| `Block`                    | character | Field block (design control for spatial effects). |
| `Week`                     | numeric   | Week number (2–14) of the time‑course harvest. |
| `Replicate`                | numeric   | Biological replicate number (1–4). |
| `Genotype`                 | character | Sorghum variety: `RT430` (stay‑green) or `BT642`. |
| `Condition`                | character | Water regime: `Preflowering` (drought before flowering), `Postflowering` (drought after flowering), or `Control` (well‑watered). |
| `Sample.type`              | character | Tissue sampled: `Leaves`. |
| `Day`                      | character | Harvest label, e.g. `Day21`. |
| `Row`                      | character | Field row. |
| `isFloweringWeek`          | logical   | `TRUE` if this sample was harvested during the plant's flowering week. |
| `Date.Harvested`           | character | Calendar harvest date (mm/dd/yy). |
| `Collection.Time`          | character | Time of day harvested. |
| `No.plants.pooled`         | numeric   | Number of plants pooled into this sample. |
| `FloweringJulianDate`      | numeric   | Julian day (day of year) when the plant flowered. |
| `FloweringDate`            | character | Calendar date when the plant flowered. |
| `FloweringWeek`            | numeric   | Week number when flowering occurred. |
| `FloweringDay`             | numeric   | Day of week when flowering occurred. |
| `Unreplicated`             | logical   | `TRUE` if this sample was not replicated. |
| `CompleteSampleTreatment`  | character | Concatenation of genotype, condition, and week, e.g. `RT430:03:Preflowering`. |
| `Timepoint`                | numeric   | Week number (same as `Week`, used for modeling). |
| `Group`                    | character | Genotype and condition combined, e.g. `RT430:Preflowering`. |

---

# Quick-reference: which files link to which {#sec-links}

```{r}
#| label: tbl-links
#| tbl-cap: "Foreign-key links between files (which column joins them)."

tribble(
  ~Dataset,                   ~Link_column,    ~Joins_to,
  # -- Yogurt --
  "01_participant_metadata_yogurt.csv",  "pid",      "00_sample_ids_yogurt.csv (pid)",
  "02_qpcr_results_yogurt.csv",         "sample_id","00_sample_ids_yogurt.csv (sample_id)",
  "03_luminex_results_yogurt.csv",      "sample_id","00_sample_ids_yogurt.csv (sample_id)",
  # -- Menstruation --
  "01_participant_metadata_period.csv", "pid",      "00_sample_ids_period.csv (pid)",
  "02_luminex_period.csv",              "sample_id","00_sample_ids_period.csv (sample_id)",
  "03_flow_cytometry_period.csv",       "sample_id","00_sample_ids_period.csv (sample_id)",
  # -- Instructional --
  "01_participant_metadata_UKZN*.csv",  "pid",      "00_sample_ids_UKZN*.csv (pid)",
  "02_visit_clinical_measurements*.csv","pid",      "00_sample_ids_UKZN*.csv (pid)",
  "03_elisa_cytokines_UKZN*.csv",      "sample_id","00_sample_ids_UKZN*.csv (sample_id)",
  "04_flow_cytometry_UKZN*.csv",       "sample_id","00_sample_ids_UKZN*.csv (sample_id)"
) |>
  knitr::kable()
```